Responsible Machine Learning\\\large{with Interpretable Models, Post-hoc Explanation, and Disparate Impact Testing}

### 2019-11-06 article-information-2019.tex
wireframe notebooks, lower-case d

%=================================================================
\documentclass[information,article,submit,moreauthors,pdftex]{Definitions/mdpi}

@@

% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0000-0000-000-000X} % Add \orcidA{} behind the author's name
%\newcommand{\orcidauthorB}{0000-0000-000-000X} % Add \orcidB{} behind the author's name

% Authors, for the paper (add full first names)
\Author{Navdeep Gill $^{1, \ddagger}$, Patrick Hall $^{1,\ddagger,*}$, Kim Montgomery $^{1,\ddagger}$, and Nicholas Schmidt $^{2, \ddagger}$}

@@

% Current address and/or shared authorship
%\firstnote{Current address: Affiliation 3} 
\secondnote{These authors contributed equally to this work.}
% The commands \thirdnote{} till \eighthnote{} are available for further notes

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This text outlines a viable approach for training and evaluating machine learning systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines (M-GBM) and explainable neural networks (XNN), a deep learning architecture well-suited for structured data, are assessed on simulated data with known feature importance and sociological bias characteristics and on realistic, publicly available lending data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence and individual conditional expectation (ICE) and global and local gradient-based or Shapley feature importance. The constrained model predictions are also tested for disparate impact and other types of sociological bias using straightforward group fairness measures. By combining innovations in interpretable models, post-hoc explanation, and bias testing with accessible software tools, this text aims to provide a template workflow for important machine learning applications that require high accuracy and interpretability and low disparate impact.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Notation}

To facilitate descriptions of data, modeling, explanatory, and social bias techniques, notation for input and output spaces, datasets, and models is defined.

@@

	\begin{itemize}
		\item A type of machine learning model $g$, selected from a hypothesis set $\mathcal{H}$, is trained to represent an unknown signal-generating function $f$ observed as  $\mathbf{X}$ with labels $\mathbf{Y}$ using a training algorithm $\mathcal{A}$: 
		$ \mathbf{X}, \mathbf{Y} \xrightarrow{\mathcal{A}} g$, such that $g \approx f$.
		\item $g$ generates learned output responses on the input dataset $g(\mathbf{X}) = \mathbf{\hat{Y}}$, and on the general input space $g(\mathcal{X}) = \mathcal{\hat{Y}}$.
		\item The model to be explained and tested for unwanted social bias is denoted as $g$.
	\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Data Description}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Model Description}
% both additive
% XNN smooth

@@

Explainable neural networks (XNNs) are an alternative formulation of additive index models in which the ridge functions are neural networks \cite{wf_xnn}. XNNs also bare a strong resemblance to generalized additive models (GAMs) and so-called explainable boosting machines (EBMs or GA\textsuperscript{2}M), i.e. GAMs which consider main effects and a small number of 2-way interactions and incorporate boosting in their training \cite{esl}, \cite{ga2m}.  Hence, XNNs enable users to tailor interpretable neural network architectures to a given prediction problem and to visualize model behavior by plotting ridge functions. XNNs are composed of a global bias term, $\mu_0$, $K$ individually specified neural networks, $n_k$ with scale parameters $\gamma_k$, and the inputs to each $n_k$ are themselves a linear combination of modeling inputs, $\sum^J_{j=0}\beta_{k, j}x_j$.

@@

Monotonic gradient boosting machines (M-GBMs) constrain typical GBM training to consider only tree splits that obey user-defined positive and negative monotonicity constraints. The M-GBM remains an additive combination of $B$ trees trained by gradient boosting, $T_b$, but each tree learns a set of splitting rules that respect monotonicity constraints,  $\Theta^\text{mono}_b$.

@@

\noindent As in unconstrained GBM, $\Theta^{\text{mono}}_b$ is selected in a greedy, additive fashion by minimizing a regularized loss function that considers known target labels, $\mathbf{y}$, the predictions of all subsequently trained trees in the M-GBM, $g^{\text{mono}}_{b-1}(\mathbf{X})$, and a regularization term that penalizes complexity in the current tree, $\Omega(T_b)$. For the $b$-th iteration, the loss function, $\mathcal{L}_{b}$, can generally be defined as:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Explanatory Method Description}

@@

Partial dependence (PD) plots are a widely-used method for describing the average predictions of a complex model $g$ across some partition of data $\mathbf{X}$ for some interesting input feature $X_j$ \cite{esl}. Individual conditional expectation (ICE) plots are a newer method that describes the local behavior of $g$ for a single instance $\mathbf{x} \in \mathcal{X}$. Partial dependence and ICE can be combined in the same plot to compensate for known weaknesses of partial dependence, to identify interactions modeled by $g$, and to create a holistic portrait of the predictions of a complex model for some $X_j$  \cite{ice_plots}.

Following \citet{esl} a single feature $X_j \in \mathbf{X}$ and its complement set $\mathbf{X}_{(-j)} \in \mathbf{X}$ (where $X_j \cup \mathbf{X}_{(-j)} = \mathbf{X}$) is considered. $\text{PD}(X_j, g)$ for a given feature $X_j$ is estimated as the average output of the learned function $g(\mathbf{X})$ when all the observations of $X_j$ are set to a constant $x \in \mathcal{X}$ and $\mathbf{X}_{(-j)}$ is left unchanged. $\text{ICE}(x_j, \mathbf{x}, g)$ for a given instance $\mathbf{x}$ and feature $x_j$ is estimated as the output of $g(\mathbf{x})$ when $x_j$ is set to a constant $x \in \mathcal{X}$ and all other features $\mathbf{x} \in \mathbf{X}_{(-j)}$ are left untouched. Partial dependence and ICE curves are usually plotted over some set of constants $x \in \mathcal{X}$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Social Bias Test Description}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Software Resources}
% Description of code

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

% richness of possible interpretable model architecture

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
% Model debugging
% Model management and monitoring
% Appeal

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\abbreviations{The following abbreviations are used in this manuscript:\\}

>>>>

%=================================================================
\documentclass[information,article,submit,moreauthors,pdftex]{definitions/mdpi}

@@

% Author Orchid ID: enter ID or remove command
%\newcommand{\orcidauthorA}{0000-0000-000-000X} % Add \orcidA{} behind the author's name
%\newcommand{\orcidauthorB}{0000-0000-000-000X} % Add \orcidB{} behind the author's name

% Authors, for the paper (add full first names)
\Author{Navdeep Gill $^{1, \dagger}$, Patrick Hall $^{1,\dagger,*}$, Kim Montgomery $^{1,\dagger}$, and Nicholas Schmidt $^{2, \dagger}$}

@@

% Current address and/or shared authorship
\firstnote{All authors contributed equally to this work.} 
%\secondnote{}
% The commands \thirdnote{} till \eighthnote{} are available for further notes

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This text outlines a viable approach for training and evaluating machine learning (ML) systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines (MGBM) and explainable neural networks (XNN), a deep learning architecture well-suited for structured data, are assessed on simulated data with known feature importance and sociological bias characteristics and on realistic, publicly available lending data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence (PD) and individual conditional expectation (ICE) and global and local gradient-based or Shapley feature importance. The constrained model predictions are also tested for disparate impact (DI) and other types of sociological bias using straightforward group fairness measures. By combining innovations in interpretable models, post-hoc explanation, and bias testing with accessible software tools, this text aims to provide a template workflow for important ML applications that require high accuracy and interpretability and low disparate impact.}

@@

ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{\href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{See: ``When a Computer Program Keeps You in Jail''}.} ML models can perpetuate and exacerbate social biases \cite{fairmlbook}, \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}.  ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}. The authors make no claim that the interdependent issues of opaqueness, unwanted sociological bias, or security vulnerabilities in ML have been solved, even as singular entities, much less as complex intersectional phenomena, e.g. the fairwashing of biased models with ML explanations or the privacy harms of ML explanations \cite{fair_washing}, \cite{shokri2019privacy}. However, this text does present an attempt to address the technological aspects of these vexing problems with interpretable models, post-hoc explanation, and DI testing implemented in widely available, free, and open source Python tools. Section \ref{sec:m_and_m} describes methods and materials herein, including simulated and collected training datasets, interpretable and constrained architectures used to train highly transparent models, post-hoc explanations used to create an \textit{appealable} decision-making framework, tests for DI and other unwanted sociological bias, and public and open source software resources. In Section \ref{sec:res}, interpretable and constrained modeling results are compared to less interpretable and unconstrained models and post-hoc explanation and DI testing results are also presented. Section \ref{sec:disc} then discusses some nuances of the presented modeling, explanation, and DI testing methods and results. Finally, Section \ref{sec:con} closes this text with a brief outline of proposed additional steps to increase human trust and understanding in ML and also touches on the authors' future work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}\label{sec:m_and_m}

Section \ref{sec:m_and_m} presents descriptions of notation, training data, ML models, post-hoc explanation techniques, DI testing methods, and software resources as follows:\\

\vspace{-10pt}
\textsection \ref{ssec:not} \textbf{Notation}: for spaces, datasets, and models.\\
\indent \textsection \ref{ssec:data} \textbf{Training data}: simulated data with known feature importance and social bias characteristics and collected home mortgage data.\\
\indent \textsection \ref{ssec:models} \textbf{ML models}: interpretable XNN and constrained MGBM models.\\
\indent \textsection \ref{ssec:expl} \textbf{Post-hoc explanation techniques}: PD, ICE, and Shapley values.\\
\indent \textsection \ref{ssec:di}  \textbf{DI testing methods}: .\\
\indent \textsection \ref{ssec:soft} \textbf{Software resources}: GitHub repository for the results presented herein and utilized and useful Python packages.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Notation}\label{ssec:not}

To facilitate descriptions of data and modeling, explanatory, and DI testing techniques, notation for input and output spaces, datasets, and models is defined here.

@@

	\begin{itemize}
		\item A type of ML model $g$, selected from a hypothesis set $\mathcal{H}$, is trained to represent an unknown signal-generating function $f$ observed as  $\mathbf{X}$ with labels $\mathbf{Y}$ using a training algorithm $\mathcal{A}$: 
		$ \mathbf{X}, \mathbf{Y} \xrightarrow{\mathcal{A}} g$, such that $g \approx f$.
		\item $g$ generates learned output responses on the input dataset $g(\mathbf{X}) = \mathbf{\hat{Y}}$, and on the general input space $g(\mathcal{X}) = \mathcal{\hat{Y}}$.
		\item The model to be explained and tested for unwanted social bias is denoted as $g$.
	\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Data Description}\label{ssec:data}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Model Description}\label{ssec:models}
% both additive
% XNN smooth

@@

XNNs are an alternative formulation of additive index models in which the ridge functions are neural networks \cite{wf_xnn}. XNNs also bare a strong resemblance to generalized additive models (GAMs) and so-called explainable boosting machines (EBMs or GA\textsuperscript{2}M), i.e. GAMs which consider main effects and a small number of 2-way interactions and incorporate boosting in their training \cite{esl}, \cite{ga2m}.  Hence, XNNs enable users to tailor interpretable neural network architectures to a given prediction problem and to visualize model behavior by plotting ridge functions. XNNs are composed of a global bias term, $\mu_0$, $K$ individually specified neural networks, $n_k$ with scale parameters $\gamma_k$, and the inputs to each $n_k$ are themselves a linear combination of modeling inputs, $\sum^J_{j=0}\beta_{k, j}x_j$.

@@

MGBMs constrain typical GBM training to consider only tree splits that obey user-defined positive and negative monotonicity constraints. The MGBM remains an additive combination of $B$ trees trained by gradient boosting, $T_b$, but each tree learns a set of splitting rules that respect monotonicity constraints,  $\Theta^\text{mono}_b$.

@@

\noindent As in unconstrained GBM, $\Theta^{\text{mono}}_b$ is selected in a greedy, additive fashion by minimizing a regularized loss function that considers known target labels, $\mathbf{y}$, the predictions of all subsequently trained trees in the MGBM, $g^{\text{mono}}_{b-1}(\mathbf{X})$, and a regularization term that penalizes complexity in the current tree, $\Omega(T_b)$. For the $b$-th iteration, the loss function, $\mathcal{L}_{b}$, can generally be defined as:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Explanatory Method Description}\label{ssec:expl}

@@

PD plots are a widely-used method for describing the average predictions of a complex model $g$ across some partition of data $\mathbf{X}$ for some interesting input feature $X_j$ \cite{esl}. ICE plots are a newer method that describes the local behavior of $g$ for a single instance $\mathbf{x} \in \mathcal{X}$. PD and ICE can be combined in the same plot to compensate for known weaknesses of PD, to identify interactions modeled by $g$, and to create a holistic portrait of the predictions of a complex model for some $X_j$  \cite{ice_plots}.

Following \citet{esl} a single feature $X_j \in \mathbf{X}$ and its complement set $\mathbf{X}_{(-j)} \in \mathbf{X}$ (where $X_j \cup \mathbf{X}_{(-j)} = \mathbf{X}$) is considered. $\text{PD}(X_j, g)$ for a given feature $X_j$ is estimated as the average output of the learned function $g(\mathbf{X})$ when all the observations of $X_j$ are set to a constant $x \in \mathcal{X}$ and $\mathbf{X}_{(-j)}$ is left unchanged. $\text{ICE}(x_j, \mathbf{x}, g)$ for a given instance $\mathbf{x}$ and feature $x_j$ is estimated as the output of $g(\mathbf{x})$ when $x_j$ is set to a constant $x \in \mathcal{X}$ and all other features $\mathbf{x} \in \mathbf{X}_{(-j)}$ are left untouched. PD and ICE curves are usually plotted over some set of constants $x \in \mathcal{X}$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Social Bias Test Description}\label{ssec:di}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Software Resources}\label{ssec:soft}
% Description of code

% github repo, main packges

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}\label{sec:res}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{sec:disc}

@@

% somewhat aligned with United States regulation-mandated adverse action notices?

%Interpretable models are also tested for disparate impact and other unwanted sociological bias and viable bias mitigation strategies are discussed. While the focus of this paper is not ML security, best-practices from that field do point to transparency of ML systems as a mitigating factor for some ML attacks and hacks \cite{papernot2018marauder}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}\label{sec:con}
% Model debugging
% Model management and monitoring
% Appeal
% local bias

% Moreover, intersectional dangers at the nexus of sociological bias, opaque, and insecure decision-making systems: privacy dangers of explanations, fairwashing, local bias, proxies
% accuracy, fairness, security are not static

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\abbreviations{The following abbreviations are used in this manuscript: ML -- machine learning, MGBM -- monotonic gradient boosting machine, XNN -- explainable neural network, PD -- partial dependence, ICE -- individual conditional expectation, DI -- disparate impact, SGD -- stochastic gradient descent.}
